Model matters with restricted Boltzmann machines

### 2019-11-16 presentations/silo_wisc/slides.Rmd
silo presentation

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan
shortname: Kaplan, et al.
institute: |
    | Duke University
    | andrea.kaplan@duke.edu
shortinstitute: andrea.kaplan@duke.edu
date: |
  | May 9, 2018
  |
  | Slides available at <http://bit.ly/kaplan-msmlc>
  | 
  | \footnotesize Joint work with D. Nordman and S. Vardeman
shortdate: "May 9, 2018"
output: 
  beamer_presentation:
    template: ../resources/latex/beamer.tex
    includes:
      in_header: ../resources/latex/front-matter.tex
theme: CambridgeUS
bibliography: ../resources/latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

@@

1. The model properties are largely unexplored in the literature
2. The commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]

@@

- All probability will stack up on mode sets or potentially those few outcomes with the highest probability 
- Proofs found in [@kaplan2016note]

@@

# Data coding to mitigate degeneracy

\vspace{-.1in}
\begin{figure}
\includegraphics[width=.55\linewidth]{../resources/images/toyhull_top.png}
\includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png}
\caption{Convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
\end{figure}
\vspace{-.15in}
- For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
- At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!

@@

# Manageable (a.k.a. small) examples

@@

- RBMs used for classification, but are concerning as statistical models due to *near-degeneracy*, *S-instability*, and *uninterpretability*
- Rigorous fitting methodology is difficult due to the dimension of the parameter space & size of the latent variable space
- Fitting a RBM model is also questionable as any distribution for the visibles can be approximated arbitrarily well
    - The empirical distribution of visibles is the best fitting model for observed cell data 
    - There can be no "smoothed distribution" achieved in a RBM model of sufficient size with a rigorous likelihood-based method

Skeptical that any model built using RBMs (i.e. deep Boltzmann machine) can achieve useful **prediction** or **inference** in a principled way without limiting the flexibility of the fitted model

# Future work

1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
\vfill
2. Approximate Bayes methods for fitting more realistic sized data (Variational Bayes) with shrinkage priors
\vfill
3. Image classification
    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
    - Decision theoretic based approach to approximating the likelihood ratio test for classification
\vfill

@@

* Slides -- <http://bit.ly/kaplan-msmlc>

* Contact
    * Email -- <andrea.kaplan@duke.edu>
    * Twitter -- <http://twitter.com/andeekaplan>
    * GitHub -- <http://github.com/andeek>

>>>>

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan
shortname: andee.kaplan@colostate.edu
institute: |
    | Colorado State University
    | andee.kaplan@colostate.edu
date: |
  | November 20, 2019
  |
  | Slides available at <http://bit.ly/kaplan-silo>
  | 
  | \footnotesize Joint work with D. Nordman and S. Vardeman
shortdate: "November 20, 2019"
output: 
  beamer_presentation:
    keep_tex: true
    template: ../resources/latex/beamer.tex
    includes:
      in_header: ../resources/latex/front-matter.tex
theme: CambridgeUS
bibliography: ../resources/latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical, @kaplan2019properties
---

@@

1. The model properties are largely unexplored in the literature
2. The commonly cited fitting methodology remains heuristic-based [@hinton2006fast]

@@

- All probability will stack up on mode sets or potentially those few outcomes with the highest probability

@@

<!-- # Data coding to mitigate degeneracy -->

<!-- \vspace{-.1in} -->
<!-- \begin{figure} -->
<!-- \includegraphics[width=.55\linewidth]{../resources/images/toyhull_top.png} -->
<!-- \includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png} -->
<!-- \caption{Convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.} -->
<!-- \end{figure} -->
<!-- \vspace{-.15in} -->
<!-- - For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space. -->
<!-- - At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*! -->

@@

# FOES models

Statements about near-degeneracy, instability, and uninterpretability hold for a general class of models that includes the RBM.

- $\boldsymbol X_N = (X_1, \dots, X_N)$ collection of discrete random variables with a finite sample space

- $\mathcal{X}^N$ finite sample space, represented as some $N$-fold Cartesian  product

- $P_{{\thetaN}}$ probability model on $\mathcal{X}^N$ for each $N$

- Assume that the model support of $P_{{\thetaN}}$ is the sample space $\mathcal{X}^N$.

#### Definition

This framework produces *Finite Outcome Everywhere Supported (FOES)* models, $P_{{\thetaN}}$, indexed by a defining sequence of parameters ${\thetaN}$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$.

[@kaplan2016note]

# Examples

- Discrete exponential family models

- Discrete restricted Boltzmann machines

- Binary deep learning models

  \vspace{.1in}

  1. **Deep Boltzmann machine (DBM).** Conditional independence within all layers, stacked RBM models with conditional dependence between neighboring layers

  \vspace{.1in}

  2. **Deep belief network (DBN).** Multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers;  all but the last stacked layer are Bayesian networks (directed dependence), rather than RBMs

# Visualizing properties in manageable (a.k.a. small) RBMs

@@

- RBMs shown success for classification, but concerning as statistical models due to *near-degeneracy*, *S-instability*, and *uninterpretability*
- Rigorous fitting methodology is difficult 
    - Numerical complications in likelihood maximization may occur due to degeneracy and S-instability when all probability in pushed to opposite extremes in the sample space 
    - For Bayesian inference, S-instability may hinder effective chain mixing
- RBMs questionably useful as any distribution for the visibles can be approximated arbitrarily well
    - The empirical distribution of visibles is the best fitting model for observed cell data 
    - There can be no "smoothed distribution" achieved in a RBM model of sufficient size with a rigorous likelihood-based method

Skeptical that any model built using RBMs (e.g., DBM) can achieve useful **inference** in principled way without limiting flexibility of the fitted model

# Relevant papers

\vfill

1. Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. "Properties and Bayesian fitting of restricted Boltzmann machines." *Statistical Analysis and Data Mining: The ASA Data Science Journal* 12.1 (2019): 23-38.

\vspace{0.25in}

2. Kaplan, Andee, Daniel J. Nordman, and Stephen B. Vardeman. "On the S-instability and degeneracy of discrete deep learning models." *Information and Inference: A Journal of the IMA* (2019).

\vfill

@@

* Slides -- <http://bit.ly/kaplan-silo>

* Contact
    * Email -- <andee.kaplan@colostate.edu>
    * Twitter -- <http://twitter.com/andeekaplan>
    * GitHub -- <http://github.com/andeek>
